---
title: "Plotting a New Index"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

April 14, 2020

How well does the time-trended COLI index track with CPIu inflation rates? 

To learn about this, we 

1) Assign each COLI a matching CPI city by examining their geometries mapped onto census-defined regions. We used a base city (Houston TX) to re-index the COLI data by year. Because the COLI indicies are relational between cities, and the sample of cities changes from year to year, the cardinal index number is meaningless by itself. It reflects the price level of city $i$ relative to that quarter's sample mean. So, if the Houston TX COLI = 100 for $t = 1$, then 150 at time $t = 2$, it is not necessarily the case that prices inflated in Houston. Then,

2) Indexed all cities (note: Houston's CPI and the COLI are set equal during this transformation) by their scaled new index then multiplied by the correpsonding CPI and divided by 100.


## Load packages and the (cleaned) COLI data
The data used here contains the COLI time-trended index and can be re-created using the RMarkdown file `coli_and_cpi_3-02.Rmd`.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(stringr)
library(magrittr)
library(here)

cpi <- read.csv("/Users/ryankobler/Desktop/coli research/rproject2/modified_data/cpi_cleaned_3-02.csv")

coliq <- read.csv("/Users/ryankobler/Desktop/coli research/rproject2/modified_data/coli_3-02.csv")

cpi$city <- as.character(cpi$city)
coliq$city <- as.character(coliq$city)

bestcities <- data.frame(table(coliq$city)) %>% arrange(desc(Freq)) %>%
  pull(Var1)
```

## Creating inflation rate variable: 2 ways
We want inflation for each time index, the CPIu index from census and our time-trended COLI index. Both should follow roughly the same trend. 

1) For both indicies $y = (y_{CPI}, y_{COLI})$, this variable is created by the following transformation: 
$$
\frac{y_{t} - y_{t-1}}{y_{t-1}} \cdot 100 = \left(\frac{y_t}{y_{t-1}} - 1\right) \cdot 100
$$

2) The three-year moving average helps smooth out noise from the time-trended index and is defined as follows:

$$
\frac{y_{t-1} + y_{t} + y_{t+1}}{3}
$$

where $y_t$ is our time-trended index at time $t$, generated by combining the CPI and COLI. This resulting variable is called `pi_new_ma3`.

```{r}
# generate CPI inflation variables
coliq %<>% 
  dplyr::group_by(city) %>%
  dplyr::mutate(pi_cpi = (CPIu/lag(CPIu) - 1)  * 100,
                pi_cpi_ma3 = (lag(pi_cpi) + pi_cpi + lead(pi_cpi)) / 3) %>%
  ungroup()

# define function that takes data frame and time-trended index, 
# returns a vector of the time-trended index.
getInflation <- function(data, variable, moving.avg = T){
  variable <- enquo(variable)
  index <- ifelse(moving.avg, "pi_new_ma3", "pi_new")
  data %>% 
  dplyr::group_by(city) %>%
  dplyr::mutate(pi_new = (!!variable/lag(!!variable) - 1)  * 100,
                pi_new_ma3 = (lag(pi_new) + pi_new + lead(pi_new)) / 3) %>%
  ungroup() %>% 
    pull(!!index)
}

coliq %<>%
  mutate(pi_new_ma3_tx = getInflation(coliq, my_index_tx),
         pi_new_ma3_ga = getInflation(coliq, my_index_ga),
         pi_new_ma3_oh = getInflation(coliq, my_index_oh),
         pi_new_ma3_mo = getInflation(coliq, my_index_mo))

```

### Sanity Check: see if the pi_old and pi from the CPI data frame are the same
This (hidden) just ensures I've properly created inflation rates.

```{r eval=FALSE, include=FALSE}
# t1 <- cpi %>% 
#   filter(city == "Houston-The Woodlands-Sugar Land" & year > 1989 & year < 2019) %>% 
#   select(year, pi) 
# t2 <- coliq2 %>% filter(city == "Houston TX") %>% select(year, pi_cpi)
# t1$year
# t2$year
# t1$pi == t2$pi_cpi
```


### Define plotting function 
```{R echo=FALSE}
# initialize time frame and which coli city we want to see
plotInflation <- function(COLI_CITY, ..., show.inflation = F, 
                          show.legend = F, 
                          moving.avg = T){
  TITLE <- ifelse(show.inflation, " Inflation", "")
  #NEW_INDEX <- ifelse(moving.avg, "pi_new_ma3", "pi_new")
  CPI <- ifelse(moving.avg, "pi_cpi_ma3", "pi_cpi")

  
  df <- coliq %>% filter(city == COLI_CITY) %>%
    select(year, CPI, ...) %>%
    pivot_longer(cols = c(CPI, ...), 
                 names_to = "infl_measure") %>%
    filter(year > 1990 & year < 2019)

  ggplot(df, aes(y = value, x = year, color = infl_measure)) + 
    geom_line(show.legend = show.legend) +
    labs(title = paste0(COLI_CITY, TITLE),
         y = "",
         x = "") + 
    
    theme_minimal()
}
```

## Draw the plots


```{r echo=FALSE, warning=FALSE, include = F, eval = F}
n <- length(bestcities)
plots <- list()
for(i in 1:n){
  plots[i] <- list(plotInflation(bestcities[i], "pi_new_ma3_tx"))
}

r <- 2
c <- 2

cowplot::plot_grid(plotlist = plots[1:(r*c)], nrow = r, ncol = c)
cowplot::plot_grid(plotlist = plots[5:8], nrow = r, ncol = c)
cowplot::plot_grid(plotlist = plots[9:12], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[13:16], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[17:20], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[21:24], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[25:28], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[29:32], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[33:36], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[37:40], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[41:44], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[45:48], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[49:52], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[53:56], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[57:60], nrow = r, ncol = c)
# cowplot::plot_grid(plotlist = plots[61:64], nrow = r, ncol = c)


```

## Comparing inflation measured with four different fixed-points
- Houston, TX 

- Atlanta, GA

- Columbia, MO

- Dayton, OH

```{r echo=FALSE, warning=FALSE}
indicies <- c("pi_new_ma3_tx", "pi_new_ma3_ga", "pi_new_ma3_mo", "pi_new_ma3_oh")

# both moving averages 3 different moving 
plotInflation("Cedar City UT", indicies,
              show.legend = T)

plotInflation("Findlay OH", indicies,
              show.legend = T)

plotInflation("Jacksonville FL", indicies,
              show.legend = T)

plotInflation("Flagstaff AZ", indicies,
              show.legend = T)

plotInflation("Jonesboro AR", indicies,
              show.legend = T)

plotInflation("Knoxville TN", indicies,
              show.legend = T)

```

### Which cities are missing the time-trend?

```{r}
missing_cities <- coliq %>%
  filter(is.na(my_index_tx)) %>%
  group_by(city) %>%
  select(CPIu, nearest_cpi, pi_cpi) %>%
  mutate(n = n()) %>%
  arrange(desc(n)) %>%
  pull(city) %>% unique
#missing_cities # there are 311 cities unaccounted for with my_index_tx and
                # there are 340 cities present with this index


```

### Plot: variance of the estimators across time

```{r}
plotVariance <- function(data, ...){
  data %>%
  group_by(year) %>%
  select(year, ...)  %>%
  mutate_all(~var(., na.rm = T)) %>%
  slice(1) %>%
  pivot_longer(cols = setdiff(colnames(.), "year"),
              names_to = "index_type") %>%
  ggplot(aes(x = year, y = value, color = index_type)) +
  geom_line() +
  labs(title = "Variance of price indicies", y = "") +
  theme_minimal()
}

plotVariance(coliq, my_index_tx, my_index_tx_hous, CPIu, COMPOSITE_INDEX) 
plotVariance(coliq, my_index_tx, my_index_tx_hous, CPIu, COMPOSITE_INDEX) 

```

### Plot: correlation between categories and the new index 

```{r}
getCorrelation <- function(data = coliq, yr, index.sm, index.big = my_index_tx){
  index.sm <- enquo(index.sm)
  index.big <- enquo(index.big)

  df <- data %>%
    filter(year == yr) %>%
    select(!!index.sm, !!index.big)
  cor(df %>% pull(!!index.sm), df %>% pull(!!index.big), use = "pairwise.complete.obs")
}


years <- 1990:2018
housing <- sapply(years, getCorrelation, data = coliq, 
                  index.sm = HOUSING, index.big = my_index_tx)
grocery <- sapply(years, getCorrelation, data = coliq, 
                  index.sm = GROCERY_ITEMS, index.big = my_index_tx)
utilities <- sapply(years, getCorrelation, data = coliq, 
                  index.sm = UTILITIES, index.big = my_index_tx)
transport <- sapply(years, getCorrelation, data = coliq, 
                  index.sm = TRANSPORTATION, index.big = my_index_tx)
health <- sapply(years, getCorrelation, data = coliq, 
                  index.sm = HEALTH_CARE, index.big = my_index_tx)
misc <- sapply(years, getCorrelation, data = coliq, 
                  index.sm = MISC_GOODS_SERVICES, index.big = my_index_tx)

corr.weights.df <- data.frame(housing = housing,
           grocery = grocery,
           utilities = utilities,
           transport = transport, 
           health = health, 
           misc = misc, 
           year = years)

corr.weights.df %>%
  pivot_longer(cols = -c(year), names_to = "index", values_to = "value") %>%
  ggplot(aes(year, value, color = index)) + geom_line() +
  labs(title = "Correlation over time between housing and time-trended index", 
       x = "", y = "") 


#
```
