---
title: "Plotting a New Index"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Date: February 29, 2020

## Load packages and the (cleaned) COLI data
This data contains our new time trended index and can be re-created using the Rmarkdown file `coli_and_cpi_3-02.Rmd`.

```{r}
library(tidyverse)
library(sf)
library(stringr)
library(magrittr)

```

## Creating inflation rate variable
This variable is created according the following transformation: 

$$
\frac{y_{t} - y_{t-1}}{y_{t-1}} \cdot 100 = \left(\frac{y_t}{y_{t-1}} - 1\right) \cdot 100
$$

```{r}
# counts of number of times each city is observed
table(cpi$city)

cpi %<>% filter(city != "N/A") %>%
  group_by(city) %>%
  mutate(pi = (CPIu/lag(CPIu) - 1)  * 100)

bestcities <- data.frame(table(coliq2$city)) %>% arrange(desc(Freq)) %>%
  pull(Var1)

coliq2 %<>% 
  dplyr::group_by(city) %>%
  dplyr::mutate(pi_new = (cpi_time_trend/lag(cpi_time_trend) - 1)  * 100,
                pi_cpi = (CPIu/lag(CPIu) - 1)  * 100) 
```

### (Check if the pi_old and pi from the CPI data frame are the same)
This just checks to ensure I've properly created inflation rates.

```{r}
t1 <- cpi %>% 
  filter(city == "Houston-The Woodlands-Sugar Land" & year > 1989 & year < 2019) %>% 
  select(year, pi) 
t2 <- coliq2 %>% filter(city == "Houston TX") %>% select(year, pi_cpi)
t1$year
t2$year
t1$pi == t2$pi_cpi
```

## Inflation 2 measures, 2 plots
```{r}
## Calculate 3-year moving averages for coli using `zoo`
coliq2 %<>%
  group_by(city) %>%
  mutate(pi_new_ma3 = zoo::rollmean(pi_new, k = 3, fill = NA))

cpi %<>% ungroup()
coliq2 %<>% ungroup()
```

### Define plotting function 
```{R}
# initialize time frame and which coli city we want to see
YEARS <- c(1989,2019)
#COLI_CITY <- "Atlanta GA"


plotInflation <- function(COLI_CITY, show.inflation = F, moving.avg = T){
  TITLE <- ifelse(show.inflation, " Inflation", "")
  NEW_INDEX <- ifelse(moving.avg, "pi_new_ma3", "pi_new")
  CPI_CITY <- coli %>%
    filter(city == COLI_CITY) %>%
    slice(1) %>%
    pull(nearest_cpi)
  #(CPI_CITY)
  
  i1 <- cpi %>% filter(city == CPI_CITY) %>%
    select(year, pi)
  
  i2 <- coliq2 %>% filter(city == COLI_CITY) %>%
    select(year, NEW_INDEX)
  
  city_slice_df <- reshape2::melt(full_join(i1, i2, by = "year"),
                                  id.var='year') %>%
    rename(Measure = variable) %>%
    filter(year > YEARS[1] & year < YEARS[2])
  #print(city_slice_df)
  
  ggplot(city_slice_df, aes(y = value, x = year, color = Measure)) + 
    geom_line(show.legend = F) +
    labs(title = paste0(COLI_CITY, TITLE),
         y = "",
         x = "") + 
    #scale_color_manual(values = c("red", "blue"),
                      # labels = c("CPI", "New index")) +
    theme_minimal()
}
# graphy <- ggplot(data = coliq2 %>% filter(city == "Spokane WA"),
#                  aes(x = YEAR, )) + 
#   geom_line(y = cpi_time_trend) +
#   labs(title = "Spokane, WA COLI index time-trended to CPI")
```

## Draw the plots

Note that the new index uses the 3-year moving average, which is computed using the zoo package. This variable generated according to the following transformation: 

$$
\frac{y_{t-1} + y_{t} + y_{t+1}}{3}
$$

where $y$ is our time-trended index generated by combining the CPI and COLI.

```{r}
n <- length(bestcities)
plots <- list()
for(i in 1:n){
  plots[i] <- list(plotInflation(bestcities[i]))
}
cowplot::plot_grid(plotlist = plots[1:12], nrow = 4, ncol = 3)
cowplot::plot_grid(plotlist = plots[13:24], nrow = 4, ncol = 3)
cowplot::plot_grid(plotlist = plots[25:36], nrow = 4, ncol = 3)
cowplot::plot_grid(plotlist = plots[37:48], nrow = 4, ncol = 3)
cowplot::plot_grid(plotlist = plots[49:60], nrow = 4, ncol = 3)
```

## Number of COLI cities matched to each CPI geography

```{r}
coliq2$nearest_cpi
```


### List of CPIs and matching COLI_CITIES

```{r}


```


